---
title: 'Exercise 3_2_1: A Simple Plot'
author: 'Dennis Abel & Jan Schwalbach'
date: 'Introduction to R for Data Analysis'
---

```{r child = "./content/config/exercises_setup.Rmd"}
```

Plotting data is fun. We also do it for data exploration (see the section on *Exploratory Data Analysis*). In these exercises, however, we just want to familiarize ourselves with some of the graphic facilities in R.

So let's start with a simple task: creating a simple scatter plot. For this purpose, we first load the *German General Social Survey - ALLBUS 2021 SUBSET*.

```{r loadgpcovid, message=FALSE, eval=TRUE}
library(dplyr)
library(haven)

allbus_2021_dvI <- 
  read_sav(
    "./data/allbus_2021/ZA5280_v2-0-1.sav"
  ) %>% 
  sjlabelled::set_na(na = c(-1:-99, 97:98))
```

Now let's plot an actual variable from the dataset. We will work with the variables `pa01` (political_orientation) and `fr03b` (children suffer under working mothers). Political orientation is a left-right self placement with a range from 1 (left) to 10 (right). `fr03b`measures respondent's opinions on the statement: "A mother who works full-time can usually have as close a relationship with her young child as a mother who does not work" and ranges from 1 (completely agree) to 4 (completely disagree).

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Create a histogram for the variable `pa01` (political_orientation) with base R.
```

```{block, opts.label = "clues"}
You simply need the `hist()` command.
```

```{r histogram, solution = TRUE}
hist(allbus_2021_dvI$pa01)
```

```{block, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}
Create a scatter plot for the variables `pa01` (political_orientation) and `fr03b` (children suffer under working mothers).
```

```{block, opts.label = "clues"}
You simply need the `plot()` command.
```

```{r scatter, solution = TRUE}
plot(allbus_2021_dvI$pa01, allbus_2021_dvI$fr03b)
```

```{block, box.title = "3 (optional)", box.body = list(fill = "white"), box.icon = "fa-star"}
Add some jitter to the scatter plot.
```

```{block, opts.label = "clues"}
You need the `jitter()` function. Have a look at `?jitter` for more details on using the function.
```

```{r jitter, solution = TRUE}
plot(
  jitter(allbus_2021_dvI$pa01, 2), 
  jitter(allbus_2021_dvI$fr03b, 2)
)
```

```{block, box.title = "4 (optional)", box.body = list(fill = "white"), box.icon = "fa-star"}
Add some color of your choice to the plot.
```

```{block, opts.label = "clues"}
Did you check out the `ColourPicker` add-in for more modern colors?
```

```{r colorscatter, solution = TRUE}
plot(
  jitter(allbus_2021_dvI$pa01, 2), 
  jitter(allbus_2021_dvI$fr03b, 2),
  col = c("#1C86EE", "#FFFFFF", "#FFFFFF")
)
```
