---
title: 'Exercise 4_1_3: Regression reporting'
author: 'Jan Schwalbach & Dennis Abel'
date: 'Introduction to R for Data Analysis'
editor_options: 
  chunk_output_type: console
---

```{r child = "./content/config/exercises_setup.Rmd"}
```

After running a few regression analyses in the previous exercises, we will now explore some options for reporting their results.

If necessary, load the data first....

```{r load}
library(dplyr)
library(haven)

allbus_2021_cda_3 <- 
  read_spss("./data/allbus_2021/ZA5280_v2-0-1.sav") %>% 
  mutate(
    trust_parliament = as.numeric(pt03),
    party = pv01,
    political_orientation = as.numeric(pa01),
    satisfaction_democracy = ifelse(as.numeric(ps03) <= 2, 1, 0)
  )
```

In addition to the `parameters` package, this time we also need the following packages: `stargazer`, `report`, and `broom`. 

```{r pkgs, eval = F}

if (!require(parameters)) install.packages("parameters")
if (!require(stargazer)) install.packages("stargazer")
if (!require(report)) install.packages("report")
if (!require(broom)) install.packages("broom")
if (!require(performance)) install.packages("performance")

```

Before we can report anything, we, of course, first need to run a regression analysis (again). We run a simple linear regression model with the trust in parliament (`trust_parliament`) as the outcome and age (`age`), sex (`sex`), German region (`eastwest`), and political orientation (`political_orientation`) as predictors. We are also interested in an interaction effect of German region and political orientation. 

```{r regres, eval = T}
reg_model <- 
  lm(
    trust_parliament ~ age + sex + eastwest*political_orientation,
    data = allbus_2021_cda_3
  )

summary(reg_model)
```

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
Using a function from a package that allows us to view model parameters, print some more interesting information on the results of our model (including confidence intervals and p-values) in a nice tabular format.
```

```{block, opts.label = "clues"}
We can use a function from the `parameters` package for printing model parameters here.
```

```{r third-exercise, solution = TRUE}
library(parameters)
model_parameters(reg_model)
```

```{block, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}
For further use in a publication, we also want to create a typical regression table using the `stargazer` package. We want the output to be in plain text format.
```

```{block, opts.label = "clues"}
If you want to, you can also specify labels for the variables in your models as arguments in the `stargazer` function.
```

```{r fourth-exercise, solution = TRUE}
library(stargazer)

stargazer(
  reg_model,
  type = "text",
  dep.var.labels = c("Trust in Parliament"),
  covariate.labels = 
    c(
      "Age", "Female (Reference: Male)", "Diverse (Reference: Male)", 
      "Region (Reference: Western Germany)", "Political Orientation",
      "Region X Political Orientation"
      )
  )
```

```{block, box.title = "3 (optional)", box.body = list(fill = "white"), box.icon = "fa-star"}
To produce custom tables and plots, we also want to store the key parameters of our model in a tidy tibble.
```

```{block, opts.label = "clues"}
There's a function in the `broom` package for that.
```

```{r fifth-exercise, solution = TRUE}
library(broom)

tidy(reg_model)
```
